Safe Transferable Regions

### 2015-11-17 paper/type-system.tex
updated section 3

The purpose of \name's region type system is to enforce the key
invariant required for memory safety, namely that an object $o_1$ in a
region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
$R_2$ is guaranteed to outlive $R_1$. Intuitively, the invariant can
be enforced by (a). tracking \emph{outlives} relationships between
various regions in the program (b). tagging the C\# type of every
object in the program with its allocation region, and (c).  ensuring
that, when a reference is created, its target object ($o_2$) is
allocated in a region that is known to be in outlives relationship
with the object containing the reference. We now formally develop this
intuition via \fbname, our explicitly typed core language (with region
types) that incorporates the features introduced in the previous
section. \fbname builds on the Featherweight Generic Java
(FGJ)~\cite{fgj} formalism, and reuses notations and various
definitions from~\cite{fgj}, such as the definition of type
well-formedness for the core (region-free) language.

@@

Fig~\ref{fig:fb-syntax} describes the syntax of \fbname (\FB). Class
types in \FB are region-annotated variants of class types in FGJ (also
called \emph{core types}). This correspondence is reflected in the
$T@\rgn$ syntax of a region type, which is the simplest form of a
region type describing an object of core type $T$ contained in a
region $\rgn$. We let $\rgn$ range over static identifiers of regions
in \FB. The only unboxed value in \FB is $\unitval$ of type \unitZ.
Rest are objects.

@@

Class definitions in \FB extend FGJ's with region parameters
($\rho$). Informally, region variables ($\rho$) are to region
identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
they can be instantiated with region identifiers, as needed. A class
in \FB is necessarily parameterized over the allocation region of its
objects. We use $\rho^a$ to denote this parameter, and often refer to
it as \emph{class}'s allocation region, although it is really the
allocation region of its objects. An Object in \FB can contain fields
referring to objects allocated in regions ($\rhobar$) other than its
own allocation region ($\rhoalloc$), provided that the former outlive
the later (i.e., $\rhobar \outlives \rho$). In such cases, the
definition of object's class needs to be parametric over allocation
regions of its fields (i.e., their classes). Furthermore, the
constraint that such regions must outlive the allocation region of the
class needs to be made explicit in the definition. For instance, a
\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
component coming from ${\rho_1}$ and second from $\rho_2$, where all
the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
$\rhoalloc$. The corresponding definition of the \C{Pair} class is as
follows(The symbol $\extends$ should be read \emph{extends}):  
\begin{codejava}[mathescape=true]
class Pair<a $\extends$ Object, b $\extends$ Object>
          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$>
           {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); 
    this.fst = fst; 
    this.snd = snd;
  }
  a@$\rho_1$ getFst() {
    return this.fst;
  }
}
\end{codejava}
To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ {
  Object<$\rgn_0$> snd = new Object<$\rgn_0$>();
  letregion $\rgn_1$ {
    Object<$\rgn_1$> fst = new Object<$\rgn_1$>();
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                  (fst,snd);
  }
}
\end{codejava}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, the region type of \C{p} would be
\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which we abbreviate
as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}\footnote{In general, we treat
$B\inang{\tbar}@\rgn$ as an abbreviation of
$B\inang{\tbar}\inang{\rbar}$}. The $\ObjZ$ class in \FB has no
fields, hence its region type is of form $\ObjZ@\rgn$.
% In general, the \C{@} notation in a
% region type of an object \C{x} highlights that
% \C{x}, and all the objects reachable from \C{x} via references are
% allocated in a single region. We say that \C{x} is \emph{contained} in
% the region. The $\ObjZ$ class in \FB contains no references to other
% objects, hence its objects are always contained in their allocation
% region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
% $\ObjZ@\rgn$), for some region $\rgn$.

>>>>

The purpose of \name's region type system is to enforce the key
invariant required for memory safety, namely that an object $o_1$ in a
region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
$R_2$ is guaranteed to outlive $R_1$. Intuitively, the invariant can
be enforced by (a). tracking \emph{outlives} relationships between
various regions in the program (b). tagging the C\# type of every
object in the program with its allocation region, and (c).  ensuring
that, when a reference is created, its target object ($o_2$) is
allocated in a region that is known to be in outlives relationship
with the object containing the reference. We now formally develop this
intuition via \fbname (\FB), our explicitly typed core language (with region
types) that incorporates the features introduced in the previous
section. \fbname builds on the Featherweight Generic Java
(FGJ)~\cite{fgj} formalism, and reuses notations and various
definitions from~\cite{fgj}, such as the definition of type
well-formedness for the core (region-free) language.

@@

Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
We refer to the class types of FGJ as \emph{core types}.
%
The following example illustrates some  key elements of \FB.
\begin{codejava}[mathescape=true]
class Pair<a $\extends$ Object, b $\extends$ Object>
          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$>
           {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); 
    this.fst = fst; 
    this.snd = snd;
  }
  a@$\rho_1$ getFst() {
    return this.fst;
  }
}
\end{codejava}

A class in \FB is parametric over zero or more type variables (as in FGJ) as well as
one or more region variables $\rho$.
(The symbol $\extends$ should be read \emph{extends}.)
% In the above example, the class \C{Pair} is parameterized over $\rho^a, \rho_1, \rho_2$.
We refer to the first region parameter, usually denoted $\rho^a$, as the \emph{allocation
region} of the class: it serves to identify the region where an instance of the class is allocated.
%
An object in \FB can contain fields referring to objects allocated in regions ($\rhobar$) other than its
own allocation region ($\rhoalloc$), provided that the former outlive
the later (i.e., $\rhobar \outlives \rho$). In such cases, the
definition of object's class needs to be parametric over allocation
regions of its fields (i.e., their classes). Furthermore, the
constraint that such regions must outlive the allocation region of the
class needs to be made explicit in the definition. 
%
For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
component coming from ${\rho_1}$ and second from $\rho_2$, as long as
$\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ {
  Object<$\rgn_0$> snd = new Object<$\rgn_0$>();
  letregion $\rgn_1$ {
    Object<$\rgn_1$> fst = new Object<$\rgn_1$>();
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                  (fst,snd);
  }
}
\end{codejava}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, the region type of \C{p} would be
\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which we abbreviate
as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}\footnote{In general, we treat
$B\inang{\tbar}@\rgn$ as an abbreviation of
$B\inang{\tbar}\inang{\rbar}$}. The $\ObjZ$ class in \FB has no
fields, hence its region type is of form $\ObjZ@\rgn$.

%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

Class types in \FB are region-annotated variants of class types in FGJ (also
called \emph{core types}). This correspondence is reflected in the
$T@\rgn$ syntax of a region type, which is the simplest form of a
region type describing an object of core type $T$ contained in a
region $\rgn$. We let $\rgn$ range over static identifiers of regions
in \FB. The only unboxed value in \FB is $\unitval$ of type \unitZ.
Rest are objects.

@@

% In general, the \C{@} notation in a
% region type of an object \C{x} highlights that
% \C{x}, and all the objects reachable from \C{x} via references are
% allocated in a single region. We say that \C{x} is \emph{contained} in
% the region. The $\ObjZ$ class in \FB contains no references to other
% objects, hence its objects are always contained in their allocation
% region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
% $\ObjZ@\rgn$), for some region $\rgn$.
